### 2020-09-16 text/article_vedrovice.Rmd
cont analysis + cleaning

<!-- view of gender in the Neolithic, archaeological context -->
@robb2017 [128] noted lack of gender archaeology debate for the
European Neolithic in comparison with other periods.
Authors present gender in the Neolihic as being
of a very different nature than gender in the later prehistory.
It is shown that gender in the Neolithic is not clearly expressed and
biological differences between sexes are not transformed into binary cultural
categories (ibid. 140-142).
Considering the Neolithic burial evidence across Europe, authors find that it
'shows very little gender distinction; when it does, it is rarely a clear-cut
binary categorization but a matter of degree and overlap, sometimes polythetically
combined with factors such as age and localness' (ibid. 138).
In response, @bickle2020 [14] argues that 'biological sex may not have defined gender
during the LBK' although the 'sexed bodies did matter'.

Here we explore the relations between burials and bodies in terms of biological
sex, age, local origin, exceptionality of grave goods and location within the
cemetery.
We focus predominantly on investigating non-random patterns in artefact
co-occurences and spatial arragement of the cemetery in relation to the
proposed *levels* of neolithic identity, i.e. sex, age and localness.
Our question is thus how are grave good assemblages related to aforementioned
categories and whether there is a consistent relationship between
these categories forming the *identity*, grave goods and spatial arrangement
of the cemetery.

@@

<!-- sample (burials, artefacts) -->
Originally, the burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during the 19th century or later
agricultural activities.
These burials are excluded from the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
anthropological determination of sex and age were corrected based on
more recent analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.
Further details on use and function of the artefacts were gathered from
@masclanslatorre2020.
Information on localness of different bodies was derived from @hrncir2020.

Each burial was originally characterized by 55 variables but only subset of
35 variables was selected for further analysis.
Some of  the original variables are omitted due to their rare occurrence,
for instance spondylus arm ring is present in a single burial only,
thus being useless in establishing relationships between pairs of burials.
Other variables are removed because they are highly dependent on each other.

The variables fall into several categories.
Firstly there are biological and anthropological characteristics derived from
retrieved human skeletal remains including sex, age simplified into age groups,
isotope ratios and body height.

Second large category of variables consists of data on artifacts presence and
absence in the burials.
These include variables noting quantities of various types of pottery and
positions of the pottery vessels within the graves,
counts of various raw materials of chipped stone tools,
presence of polished stone axes or adzes, grinding slabs, pebbles,
bone tools and different kinds of personal ornaments.
The personal ornaments include beads made of spondylus shell or marble,
sponylus pendants or oval shaped buckles.
Presence of ochre pigment is also noted.
All the variables in this category are simplified by binarization (presence vs
absence of a trait) in order to minimize effects of only several numerous
cases on the overall pattern, for instance 20 beads in a grave <!-- 70 -->
would have much larger effect on the analysis then 1 pottery vessel in the same grave.

Last category describes physical characteristics of the grave pit and the
position of the body.
I. e. size of the pit, its orientation, position of the body on one of
the sides and direction of the head.

@@

@@

@sosna2013

To determine non-randomness of various artefact combinations we conduct a
permutation test on the presence/absence data matrix.
The method based on generalized Monte Carlo tests was first
introduced to archaeology by @manly1996 and adapted by @sosna2008.
Our approach is based on both of these works.
Original (observed) co-occurence matrix is permutated $n$ times with constant
row and column totals, i. e. both the count of artifacts in individual graves
and sum of the given type of artifacts across the burials are fixed.

To examine which associations of variables are infrequent or even not present
and which tend to appear together co-occurences of observed and simulated data
are summarized using two statistics suggested by @manly1996.
The $S$ statistic is a measure of overall deviation between expected and observed
number of artefact occurences while
the $v$ statistic is a measure of difference between co-occurences of
individual artefact with other artifacts
[@manly1996, 475, for details see SI].

With $n = 999$ permutations of the original data matrix the results obtained
are robust

>>>>

<!-- view of gender in the Neolithic, archaeological context -->
@robb2017 [128] noted lack of gender archaeology debate for the
European Neolithic in comparison with other periods.
Authors present gender in the Neolihic as being
of a very different nature than gender in the later prehistory.
Gender in the Neolithic is not clearly expressed and
biological differences between sexes are not transformed into binary cultural
categories (ibid. 140-142).
Considering the Neolithic burial evidence across Europe, authors find that it
'shows very little gender distinction; when it does, it is rarely a clear-cut
binary categorization but a matter of degree and overlap, sometimes polythetically
combined with factors such as age and localness' (ibid. 138).
In response, @bickle2020 [14] argues that 'biological sex may not have defined gender
during the LBK' although the 'sexed bodies did matter'.

Here we explore the relations between burials and bodies in terms of biological
sex, age, origin, exceptionality of grave goods and location within the
cemetery.
We focus predominantly on investigating non-random patterns in artefact
co-occurences and spatial arragement of the cemetery in relation to the
proposed *levels* of neolithic identity, i.e. sex, age and localness.
Our question is thus how are grave good assemblages related to aforementioned
categories and whether there is a consistent relationship between
these categories forming the *identity*, grave goods and spatial arrangement
of the cemetery.

@@

<!-- sample (burials, artefacts) -->
Originally, the burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during the 19th century or later
agricultural activities [@podborsky2002].
These burials are excluded from most of the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
anthropological determination of sex and age were corrected based on
more recent analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.
Further details on use and function of the artefacts were gathered from
@masclanslatorre2020.
Information on localness of different bodies was adapted from @hrncir2020.

Each burial was originally characterized by 55 variables but only subset of
35 variables was selected for further analysis.
Some of  the original variables are omitted due to their rare occurrence,
for instance spondylus arm ring is present in a single burial only,
thus being useless in establishing relationships between pairs of burials.
Some other variables are removed because they are highly interdependent.

The variables fall into several categories.
Firstly there are biological and anthropological variables derived from
retrieved human skeletal remains including sex, age simplified into age groups,
various isotope ratios and body height.

Second large category of variables consists of data on artifacts present in the burials.
These include variables noting quantities of various types of pottery and
positions of the pottery vessels within the graves,
counts of various raw materials of chipped stone tools,
presence of polished and beveled stone axes or adzes, grinding slabs, pebbles,
bone tools and different kinds of personal ornaments.
The personal ornament includes beads made of spondylus shell or marble,
sponylus pendants or oval shaped buckles.
Presence of ochre pigment in the grave is also noted.

Last category describes physical characteristics of the grave pit and the
position of the body.
I. e. size of the pit, its orientation, position of the body on one of
the sides and direction of the body and head.

Quantities of various *grave goods* are simplified into a binary matrix indicating
presence or absence of a given trait. 
Only grave goods occuring in more than 5% of the graves are used in the analysis.

@@

To determine non-randomness of various artefact combinations we conduct a
permutation test on the presence/absence data matrix.
The method based on generalized Monte Carlo tests was first
introduced to archaeology by @manly1996 and adapted by @sosna2008.
Our approach is based on both of these works.
Original (observed) co-occurence matrix is permutated $n$ times with constant
row and column totals, i.e. both the count of artifacts in individual graves
and sum of the given type of artifacts across the burials are fixed.

To examine which associations of variables, i.e. artefacts, are infrequent or even not present
and which tend to appear together more often than it would be expected by random chance only
the co-occurences of observed and simulated data
are summarized using two statistics suggested by @manly1996.
The $S$ statistic is a measure of overall deviation between expected and observed
number of artefact occurences while
the $v$ statistic is a measure of difference between co-occurences of
individual artefact with other artifacts
[@manly1996, 475, for details see SI].
To assess relations between groups based on sex, age local origin and exceptionality,
statistics introduced by @sosna2008 are used.

With $n = 999$ permutations of the original data matrix the results obtained
are robust+

@@
